Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array

### 2023-10-27 docs/manuscript/paper.tex
try some new PP runs

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman PTA]{Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array}

@@

% Abstract of the paper
\begin{abstract}	
Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing arrays. A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. A two-step procedure is applied within a state-space framework, such that the modulation is tracked with a Kalman filter, which then provides a likelihood for nested sampling. The procedure enables the static parameters in the problem, such as the sky position of the source, to be inferred without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional parameter estimation methods. It also returns the Bayes factor relating a model with a single-source gravitational wave to one without, complementing traditional detection methods. It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$. Full posterior distributions of model parameters are recovered and tested for accuracy. There is a bias of $\sim 0.3$ rad in the estimated marginalised 1D posterior for the inclination $\iota$, introduced by dropping the so-called `pulsar terms'. Smaller biases $\lesssim 1 \%$ are also observed in some other static parameters.
\end{abstract}

@@

The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. After multiple non-detections \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A} consilient evidence for the GW background was presented by NANOGrav \citep{2023ApJ...951L...8A}, EPTA/InPTA \citep{2023arXiv230616214A}, PPTA \citep{2023ApJ...951L...6R} and the CPTA \citep{2023RAA....23g5024X}. \newline

Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the early stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. Individual SMBHBs are continuous wave sources; they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals. Consequently, they are detected more efficiently by either a frequentist matched filter, e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA}, or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A}, rather than by cross-correlating pulsar pairs. However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}. Inconclusive evidence at low significance was presented recently by the EPTA for an individual source at 4-5 nHz \citep{2023arXiv230616226A}. \newline

Intrinsic pulsar timing noise -- i.e.  random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Myers2021MNRAS.502.3113M,Meyers2021,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red). In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

@@

This paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW. In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence. In Section \ref{sec:testing} we describe how we create synthetic validation data on which to test the method. We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant parameter space. In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates. Discussion on  future extensions and conclusions are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively. The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization which is deferred to future work. Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

@@

\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean, white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}

@@

In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:

@@

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth. Specifically, the pulse frequency at the Earth is modulated harmonically at the GW frequency. We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section. The measurement equation is a key input into the Kalman filter in Section \ref{sec:kalman_filter}
\subsubsection{Plane GW perturbation}\label{sec:plane_gw}
We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
\end{equation}
with nearly Lorentz spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$ \citep{schutz2022}. The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$. Throughout this paper we work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB). We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. \newline

In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic. Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) changes during the inspiral by \citep[e.g.][]{Zhu10}
\begin{equation}
	\Delta f_{\rm gw} = 3.94 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^9 M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 10^{-7} \mathrm{~Hz}]^{11/3} (T_{\rm obs}^2/10 \mathrm{yr}) \gtrsim 8 \mathrm{yr}$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$. After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames. Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{equation}
	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
	\label{eq:g_func}
\end{equation}
and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column). The top panels show the true pulsar state $f_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$  (green curve). We subtract $f_{\rm em}^{(1)}(t)$ to better illustrate the stochastic wandering of the pulsar frequency. In the left column, the blue/green solutions overlap almost perfectly; in the right column, they do not. The middle panels show the true measured frequency $f_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve). Again we subtract $f_{\rm em}^{(1)}(t)$. In the left column the red/magenta solutions overlap almost perfectly; in the right column, they do not. The bottom panels show the residual or the innovation $\epsilon(t) =f_{\rm m}^{(1)}(t) - \hat{f}_{\rm m}^{(1)}(t)$. The results are shown for a single pulsar. In the right column $\Omega$ is displaced from its true value by 20\%.} 
	\label{fig:kalman_example}
\end{figure*}

@@

\subsection{Summary of workflow}\label{sec:methodsummary}
For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
\begin{enumerate}[leftmargin=2em]
	\item Specify a PTA composed of $N$ pulsars 
	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$. 
	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
	 \item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
	 \item For each live point:
\begin{enumerate}[leftmargin=2em]
	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter.
	\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
\end{enumerate}
	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where  $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a) -- (vi)(b).
	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
	\item Repeat steps (vii) - (ix) until convergence criteria are satisfied.
\end{enumerate}
In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$. The resulting $\mathcal{Z}$ values can then be compared. We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs. The generalization to TOAs is subtle and will be tackled in a forthcoming paper.

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman PTA]{Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array}

@@

% Abstract of the paper
\begin{abstract}	
Continuous nanohertz gravitational waves from individual supermassive black hole binaries may be detectable with pulsar timing arrays. A novel search strategy is developed, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. A two-step procedure is applied within a state-space framework, such that the modulation is tracked with a Kalman filter, which then provides a likelihood for nested sampling. The procedure enables the static parameters in the problem, such as the sky position of the source, to be inferred without fitting for ensemble-averaged statistics such as the power spectral density of the timing noise, and therefore complements traditional parameter estimation methods. It also returns the Bayes factor relating a model with a single-source gravitational wave to one without, complementing traditional detection methods. It is shown via astrophysically representative software injections in Gaussian measurement noise that the procedure distinguishes a gravitational wave from pure noise down to a characteristic wave strain of $h_0 \sim 4 \times 10^{-15}$. Full posterior distributions of model parameters are recovered and tested for accuracy. There is a bias of $\sim 0.3$ rad in the marginalised one-dimensional posterior for the orbital inclination $\iota$, introduced by dropping the so-called `pulsar terms'. Smaller biases $\lesssim 1 \%$ are also observed in other static parameters.
\end{abstract}

@@

The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. After multiple non-detections \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A} consilient evidence for the GW background was presented by NANOGrav \citep{2023ApJ...951L...8A}, EPTA/InPTA \citep{2023arXiv230616214A}, PPTA \citep{2023ApJ...951L...6R} and the CPTA \citep{2023RAA....23g5024X}. \newline

Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the early stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. Individual SMBHBs are continuous wave sources; they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals. Consequently, they are detected more efficiently by either a frequentist matched filter, e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA}, or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A}, rather than by cross-correlating pulsar pairs. However, PTA observational campaigns to detect individual sources have been unsuccessful so far \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}. Inconclusive evidence at low significance was presented recently by the EPTA for an individual source at 4--5 nHz \citep{2023arXiv230616226A}. \newline

Intrinsic pulsar timing noise -- i.e.  random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Myers2021MNRAS.502.3113M,Meyers2021,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. In modern Bayesian PTA searches, the power spectral density of the red intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the noise induced by a stochastic GW background (whose spectrum is also red). In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

@@

The paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW. In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence. In Section \ref{sec:testing} we describe how we create synthetic validation data with which to test the method. We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant domain of SMBHB source parameters. In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates. Conclusions are drawn in Section \ref{sec:discussion}. The data are formulated as pulse frequency time series with Gaussian measurement noise as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization which is deferred to future work. Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

@@

\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean, white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},

@@

In Equation \eqref{eq:frequency_evolution}, $f_{\rm em}^{(n)}$ is the solution of the electromagnetic spin-down equation given by Equation \eqref{eq:spinevol}, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies

@@

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth. Specifically, the pulse frequency at the Earth is modulated harmonically at the GW frequency. We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section. The measurement equation is a key input into the Kalman filter in Section \ref{sec:kalman_filter}
\subsubsection{Plane GW perturbation}\label{sec:plane_gw}
We consider a gravitational plane wave from a single, distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
\end{equation}
with nearly Lorentz spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$ \citep{schutz2022}. The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$. Throughout this paper we work with pulsar TOAs defined relative to the Solar System barycentre (SSB). We are free to choose coordinates such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. \newline

In this paper the source is approximated as monochromatic. In reality the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) increases during the inspiral by an amount \citep[e.g.][]{Sesana2010}
\begin{equation}
	\Delta f_{\rm gw} = 0.05 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^{8.5} M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{50 \mathrm{~nHz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
%
%\begin{equation}
%	\Delta f_{\rm gw} = 3.94 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^9 M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{10^{-7} \mathrm{~Hz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
%	\label{eq:f_evolution}
%\end{equation}
where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \gtrsim 20$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression $p_{\alpha} u^{\alpha}$. After barycentering, one has $u^{\alpha} =(1,0,0,0)$ for both the emitter and the observer to leading order in the respective momentarily comoving reference frames. Hence Equation \eqref{eq:momentum_shift} can be written for the $n$-th pulsar as
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{equation}
	g^{(n)}(t) = 1 - \frac{h_{ij} (t; \boldsymbol{x}= 0)[q^{(n)}]^i [q^{(n)}]^j}{2[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] }  \left \{1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right \} \ ,
	\label{eq:g_func}
\end{equation}
and $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}  define a non-linear measurement equation that relates the intrinsic pulsar spin frequency to the pulse frequency measured by an observer on Earth.

@@

\begin{figure*}
	\includegraphics[width=\textwidth, height =0.5\textwidth]{images/Kalman_example_both}
	\caption{Sample output of the Kalman filter illustrating the accuracy of the reconstructed state sequence $f_{\rm p}^{(1)}(t)$ when the static parameters are correct (${\boldsymbol{\hat\theta}} = {\boldsymbol{\theta}}$, left column) and incorrect (${\boldsymbol{\hat\theta}} \neq {\boldsymbol{\theta}}$, right column). The top panels show the true pulsar state $f_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (blue curve) and the state estimated by the Kalman filter $\hat{f}_{\rm p}^{(1)}(t) - f_{\rm em}^{(1)}(t)$  (green curve). We subtract $f_{\rm em}^{(1)}(t)$ to better illustrate the stochastic wandering of the pulsar frequency. In the left column, the blue/green solutions overlap almost perfectly; in the right column, they do not. The middle panels show the true measured frequency $f_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (red curve) and the frequency estimated  by the Kalman filter $\hat{f}_{\rm m}^{(1)}(t) - f_{\rm em}^{(1)}(t)$ (magenta curve). Again we subtract $f_{\rm em}^{(1)}(t)$. In the left column the red/magenta solutions overlap almost perfectly; in the right column, they do not. The bottom panels show the residual or innovation $\epsilon(t) =f_{\rm m}^{(1)}(t) - \hat{f}_{\rm m}^{(1)}(t)$. In the right column $\Omega$ is displaced from its true value by 20\%. Results are shown for a single pulsar. } 
	\label{fig:kalman_example}
\end{figure*}

@@

\subsection{Summary of workflow}\label{sec:methodsummary}
For the reader's convenience we now summarise the workflow for a representative PTA analysis using the Kalman filter and nested sampler for parameter estimation and model selection:
\begin{enumerate}[leftmargin=2em]
	\item Specify a PTA composed of $N$ pulsars 
	\item Obtain $N$ data inputs $f_{\rm m}^{(n)}(t)$, collectively labelled $\boldsymbol{Y}$
	\item Specify a state-space model $\mathcal{M}$, with static parameters $\boldsymbol{\theta}$
	\item Specify prior distribution $\pi(\boldsymbol{\theta})$
	 \item Sample $n_{\rm live}$ points from $\pi(\boldsymbol{\theta})$ 
	 \item For each live point:
\begin{enumerate}[leftmargin=2em]
	\item Pass the sample $\boldsymbol{\theta}_{\rm sample}$ to the Kalman filter
	\item Iterate over the input data using the Kalman filter and obtain a single $\log \mathcal{L}$ value, Equation \eqref{eq:likelihood}
\end{enumerate}
	\item Remove the live point with the lowest likelihood value, $\log \mathcal{L}_{\rm lowest}$
	\item Sample a new live point from $\pi(\boldsymbol{\theta})$, subject to the requirement that the new likelihood obeys $\mathcal{L}_{\rm new}$ > $\mathcal{L}_{\rm lowest}$, where $\log \mathcal{L}_{\rm new}$ is calculated via steps (vi)(a)--(vi)(b).
	\item Update $p\left(\boldsymbol{\theta}|\boldsymbol{Y}\right)$ and $\mathcal{Z}$ with nested sampler
	\item Repeat steps (vii)--(ix) until convergence criteria are satisfied.
\end{enumerate}
In order to compute $\beta$ the above workflow is repeated for a different $\mathcal{M}$. The resulting $\mathcal{Z}$ values can then be compared. We remind the reader that the above workflow differs from a realistic PTA analysis in one important respect, namely that the data are input as frequency time series $f_{\rm m}^{(n)}(t)$ instead of pulse TOAs. The generalization to TOAs is subtle and will be tackled in a forthcoming paper.

\section{Validation with synthetic data} \label{sec:testing}
In this section we outline how synthetic data are generated in order to validate the analysis scheme in Section \ref{sec:detect}. Synthetic data enable validation to occur systematically and under controlled conditions. In Section \ref{sec:synt_pta} we describe how to construct a representative synthetic PTA, and how to set astronomically reasonable values for the static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$. In Section \ref{sec:gendata} we demonstrate how to solve Equations \eqref{eq:spinevol}--\eqref{eq:frequency_evolution}, and \eqref{eq:measurement}--\eqref{eq:g_func} for the synthetic PTA so as to generate noisy, frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t) $.
